### 2018-04-02 chapter4/ms2.tex
Fix Chapter 4. Ch 6 done. Begin Intro

%UNLIKE IN A REGULAR TEX FILE, DON'T PUT ANY PREAMBLE MATERIAL HERE

\section*{Abstract}
  We report the results of a Dark Energy Camera (DECam)
  optical follow-up of the gravitational wave (GW) event GW151226,
  discovered by the Advanced LIGO detectors. Our observations cover 28.8
  deg$^2$ of the localization region in the $i$ and $z$ bands (containing 3\%
  of the {\tt BAYESTAR} localization probability), starting 10 hours after the
  event was announced and spanning four epochs at $2-24$ days after the
  GW detection. We achieve $5\sigma$ point-source limiting
  magnitudes of $i\approx21.7$ and $z\approx21.5$, with a scatter
  of $0.4$~mag, in our difference images. Given the two day delay,
  we search this area for a rapidly declining optical counterpart with
  $\gtrsim 3\sigma$ significance steady decline between the first and final
  observations. We recover four sources that pass our selection criteria,
  of which three are cataloged AGN. The fourth source is offset
  by $5.8$ arcsec from the center of a galaxy at a distance of 187
  Mpc, exhibits a rapid decline by $0.5$ mag over $4$ days,
  and has a red color of $i-z\approx 0.3$ mag. These properties roughly match
  the expectations for a kilonova. However, this source
  was detected several times, starting $94$ days prior to GW151226,
  in the Pan-STARRS Survey for Transients (dubbed as PS15cdi) and
  is therefore unrelated to the GW event. Given its long-term
  behavior, PS15cdi is likely a Type IIP supernova that
  transitioned out of its plateau phase during our
  observations, mimicking a kilonova-like behavior. We comment on
  the implications of this detection for contamination in future
  optical follow-up observations.

@@

The Advanced Laser Interferometer Gravitational-Wave Observatory
(LIGO) is designed to detect the final inspiral and
merger of compact object binaries comprised of neutron stars (NS)
and/or stellar-mass black holes (BH) \citep{LIGOMainRef}.
The first LIGO observing run (designated O1)
commenced on 18 September 2015 with the ability to detect binary
neutron star (BNS) mergers to an average distance of $\sim 80$ Mpc, a
forty-fold increase in volume relative to the previous generation of
ground-based GW detectors. On 2015 September 14 LIGO detected
the first GW event ever observed, GW150914 \citep{LIGOGW150914}.

The waveform of GW150914 was consistent with the inspiral, merger,
and ring-down of a binary black hole (BBH) system ($36+29$ M$_\odot$;
\citealt{LIGOGW150914}) providing the first observational
evidence that such systems exist and merge. While there are no robust
theoretical predictions for the expected electromagnetic (EM)
counterparts of such a merger, more than 20 teams conducted a wide
range of follow-up observations spanning from radio to $\gamma$-rays, along with neutrino follow up
\citep{LIGOGW150914FollowUp,LIGOGW150914FollowUpSupp,GW150914IceCube,Annis+16,GW150914Fermi,Evans+16,Kasliwal+16,Savchenko+16,GW150914PS1,GW150914DECam,Tavani+16}.
This effort included deep optical follow-up observations by our group using DECam covering 100 deg$^2$
(corresponding to 38\%~(11\%) of the initial~(final) sky maps)
 -- making this one of the most comprehensive optical
follow-up campaigns for GW150914 \citep{Annis+16,GW150914DECam}. Our search for rapidly
declining transients to limiting magnitudes of $i\approx21.5$ mag
for red $(i-z=1)$ and $i\approx20.1$ mag for blue $(i-z=-1)$ events yielded no counterpart to
GW150914 \citep{GW150914DECam}. One result of the broader multi-wavelength
follow-up campaign is a claimed coincident detection of a weak short
gamma-ray burst (SGRB) from the {\it Fermi}/GBM detector 0.4 s after
the GW event \citep{GW150914Fermi}. However, this event was not
detected in INTEGRAL $\gamma$-ray data \citep{Savchenko+16} and was also
disputed in a re-analysis of the GBM data \citep{Greiner+16}.

A second high-significance GW event, designated GW151226, was
discovered by LIGO on 2015 December 26 at 03:38:53 UT \citep{LIGOGW151226}.
This event was also due to the inspiral and merger of a BBH system, consisting
of $14.2^{+8.3}_{-3.7}$~M$_\odot$ and $7.5^{+2.3}_{-2.3}$~M$_\odot$ black holes at a luminosity
distance of $d_L = 440^{+180}_{-190}$~Mpc \citep{LIGOGW151226}.
The initial localization was provided as a probability sky map via a private
GCN circular 37 hours after the GW detection \citep{SingerPrice16}. We initiated
optical follow-up observations with DECam 10 hours later on 2015 December 28,
and imaged a 28.8 deg$^2$ region in the $i$ and $z$ bands during several epochs. Here we report
the results of this search. In \cref{sec:ch4_obs} we discuss the
observations and data analysis procedures. In \cref{sec:ch4_analysis}
we present our search methodology for potential counterparts to GW151226,
and the results of this search. We summarize our conclusions in \cref{sec:ch4_conc}. We perform
cosmological calculations assuming $H_0 = 67.8$~km  s$^{-1}$ Mpc$^{-1}$,
$\Omega_{\lambda} = 0.69$, and $\Omega_m = 0.31$ \citep{Planck2016}.
 Magnitudes are reported in the AB system.

@@

GW151226 was detected on 2015 December 26 at 03:38:53 UT by a
Compact Binary Coalescence (CBC) search pipeline \citep{LIGOGW151226}. The CBC
pipeline operates by matching the strain data against waveform
templates and is sensitive to mergers containing NS and/or BH.
The initial sky map was generated by the {\tt BAYESTAR} algorithm
and released 37 hours after the GW detection. {\tt BAYESTAR} is a Bayesian
algorithm that generates a localization sky map based on the parameter
estimation from the CBC pipeline \citep{Singer+14,SingerPrice16}. The sky area contained within the
initial 50\% and 90\% contours was 430 deg$^2$ and 1340 deg$^2$,
respectively. A sky map generated by the {\tt LALInference}
algorithm, which computes the localization using Bayesian forward-modeling
of the signal morphology \citep{Veitch+15}, was released on 2016 January 15 UT,
after our DECam observations had been concluded. The {\tt LALInference} sky map
is slightly narrower than the sky map from {\tt BAYESTAR} with 50\% and 90\% contours
of 362 deg$^2$ and 1238 deg$^2$, respectively.

We initiated follow-up observations with DECam on 2015 December 28 UT,
two days after the GW detection and 10 hours after distribution of the {\tt BAYESTAR}
sky map. DECam is a wide-field optical imager with a 3 deg$^2$ field of view
\citep{Flaugher+15}. We imaged a 28.8 deg$^2$ region corresponding to 3\%
of the sky localization probability when convolved with the initial {\tt BAYESTAR} map
and 2\% of the localization probability in the final {\tt LALInference} sky map.
The pointings and ordering of the DECam observations were determined using
the automated algorithm described in \citet{GW150914DECam}. The choice
of observing fields was constrained by weather, instrument availability, and the available
time to observe this sky region given its high airmass. We
obtained four epochs of data with each epoch consisting of one 90 s
exposure in $i$-band and two 90 s exposures in $z$-band for each of the 12 pointings. The first
epoch was obtained 2--3 days after the initial LIGO trigger (2015
December 28--29 UT), the second epoch was at 6 days (2016 January 1
UT), the third epoch was at 13--14 days (2016 January 8--9), and the
fourth epoch was at 23--24 days (2016 January 18--19). A summary of
the observations is provided in \cref{ch4_datatable} and a visual
representation of the sky region is shown in
\Cref{fig:ch4_obs}.

\begin{deluxetable}{lrccccccc}
\singlespace
\tabletypesize{\footnotesize}
\tablecolumns{9}
\tablewidth{0pt}
\tablecaption{Summary of DECam Observations of GW151226
\label{ch4_datatable}}
\tablehead{
\colhead{Visit}    &
\colhead{UT}       &
\colhead{$\Delta t$\tablenotemark{a}} &
\colhead{$\langle$PSF$_i$$\rangle$}    &
\colhead{$\langle$PSF$_z$$\rangle$}    &
\colhead{$\langle$airmass$\rangle$}  &
\colhead{$\langle$depth$_i\rangle$}   &
\colhead{$\langle$depth$_z\rangle$}  &
\colhead{$A_{\mathrm{eff}}$\tablenotemark{b}} \\
\colhead{ } &
\colhead{ } &
\colhead{(days)} &
\colhead{(arcsec)} &
\colhead{(arcsec)} &
\colhead{ } &
\colhead{(mag)} &
\colhead{(mag)} &
\colhead{(deg$^2$)}
}
\startdata
 Epoch 1 & 2015-12-28.11 & 1.96 & 0.97 & 0.99 & 1.95 & 22.39 & 22.23 & 14.4 \\      %  6 hexes
               & 2015-12-29.11 & 2.96 & 1.00 & 0.97 & 1.78 & 22.57 & 22.46 & 14.4 \\      %  6 hexes

@@

\begin{figure}[t!]
\centering
\vspace{-60pt}
\includegraphics[width=0.75 \textwidth]{./figs/chapter4/fig1.pdf}
\vspace{-60pt}
\caption{\singlespace 
Sky region covered by our DECam observations (red
hexagons) relative to the 50\% and 90\% probability regions
from the {\tt BAYESTAR} (cyan contours) and {\tt LALInference}
(white contours) localization of GW151226.
The background color indicates the estimated $5\sigma$
point-source limiting magnitude for a 90~s $i$-band exposure
as a function of sky position for the first night of our DECam
observations. The variation in the limiting magnitude is largely
driven by the dust extinction and airmass at that position.
The dark grey regions indicate sky positions that
were unobservable due to the telescope pointing limits.  The yellow contour
indicates the region of sky covered by the Dark Energy Survey
(DES). The total effective area for the 12 DECam pointings is 28.8
deg$^2$, corresponding to 3\%~(2\%) of the probability in the
{\tt BAYESTAR}~({\tt LALInference}) sky map.}
\label{fig:ch4_obs}
\end{figure}

@@

\item We require non-negative or consistent with zero
(i.e., within 2$\sigma$ of zero)
$i$- and $z$-band fluxes in the difference photometry
across all epochs to eliminate any sources that re-brighten in the
template epoch. This provides an initial sample of 602 candidates.

@@

\clearpage
\item We require a $\ge 3\sigma$ decline in flux between the first and
third epochs to search for significant fading\footnote{\singlespace We note that this
criterion effectively requires the detection in the first epoch to be $\gtrsim 5\sigma$
producing an effectively shallower transient search. Soares-Santos et al.
\citeyearpar{GW150914DECam} quantified this effect by injecting fake sources into
their observations to determine the recovery efficiency and loss of detection
depth from analysis cuts. Here, we forego
such analysis to focus the discussion on the effects of contamination in
optical follow-up of GW events.}. We calculate $\sigma$ as the
quadrature sum of the flux errors from the first and third epochs
($\sigma = \sqrt{\sigma_1^2 + \sigma_3^2}$, where $\sigma_1$ and
$\sigma_3$ are the flux errors from the first and third epochs,
respectively). This criterion leaves a sample of 48 objects.

@@

\item We reject sources that are present as an isolated point source in
the fourth (template) epoch to eliminate variable stars. This step involved
visual inspection of the 32 candidates from step 4.

\end{enumerate}

Only four events passed our final criterion. We find that two of those events are
coincident with the nuclei of known AGN (PKS\,0129-066 and Mrk\,584), indicating
that they represent AGN variability. A third candidate is coincident with the nucleus of
the bright radio source PMN\,J0203+0956 ($F_\nu(365\,{\rm MHz})\approx 0.4$ Jy,
\citealt{Douglas+96}), also suggesting AGN variability.

\clearpage
The final candidate in our search is located at RA = 01$^{\rm h}$42$^{\rm m}$16$^{\rm s}$.17 and DEC = $-$02\arcdeg13\arcmin42 6\arcsec~(J2000), with an offset of $5.8$ arcsec from the galaxy CGCG 386-030 (RA =  01$^{\rm h}$42$^{\rm m}$15$^{\rm s}$.6, DEC = $-$02\arcdeg13\arcmin38 5\arcsec; J2000), at $z = 0.041$ or $d_L \approx 187$ Mpc (6dFGS, \citealt{Jones+04,Jones+09}); see \Cref{fig:ch4_PS15cdi}. We note that this distance is inconsistent with the 90\% confidence interval for the distance to GW151226 based on the GW data \citep{LIGOGW151226}. We observe this source in a state of rapid decline with an absolute magnitude of $M_i \approx -15$~mag on 2015 December 28 and $M_i\approx -14.5$~mag on 2016 January 1, indicating a decline rate of $\approx 0.12$ mag d$^{-1}$; the decline rate in $z$-band is $\approx 0.10$ mag d$^{-1}$. Additionally, the source exhibits a red $i-z$ color of $0.3$ mag.

@@

 \begin{figure}[t!]
 \centering
 \includegraphics[width=0.9\textwidth]{./figs/chapter4/fig2.pdf}
 \caption{\singlespace {\it Top}: Single-epoch images of our main candidate
 from all four epochs (green circle). This is the event discovered as PS15cdi in
 the PSST about 94 d prior to GW151226. {\it Bottom}: Light
 curve data for PS15cdi from PSST $w$- and $i$-band
 observations (green squares and yellow diamonds, respectively).
 Our DECam $i$- and $z$-band data are shown as
 blue circles and red stars, respectively. The revised DECam analysis
 using pre-existing templates is shown as open symbols. Upper
 limits are indicated by triangles. The inset focuses on our
 DECam data, indicating a rapid decline in both $i$ and $z$
 bands.  We fit a power-law model to the data finding a temporal index of
 $\alpha = -0.43$ (dashed-dot line). Kilonova models from Barnes and
 Kasen \citeyearpar{BarnesKasen13} with $v_{\rm ej}=0.2c$ and $M_{\rm ej}=0.1$ M$_\odot$
 at a distance of $187$~Mpc are also shown (dashed line).}
 \label{fig:ch4_PS15cdi}
 \end{figure}

\clearpage
This source was previously detected as PS15cdi on 2015 September 23 by the Pan-STARRS
Survey for Transients (PSST\footnote{\singlespace \url{http://psweb.mp.qub.ac.uk/ps1threepi/psdb/candidate/1014216170021342600/}},
\citealt{Huber+15}); see \Cref{fig:ch4_PS15cdi}. The absolute $i$-band
magnitude in the first PSST epoch, $M_i\approx -16.6$ mag and the shallow decline of $\approx0.6$~mag
over $\approx70$~d, are consistent with a Type IIP core-collapse supernova (SN).
A likely interpretation of the rapid decline in our observations is that PS15cdi is a Type IIP SN undergoing
the rapid transition from the hydrogen recombination driven plateau to the radioactive $^{56}$Co dominated
phase \citep{KasenWoosley09,Sanders+15,Dhungana+16}. The red $i-z$ color in our data is consistent with observations
of other IIP SN during this phase of evolution (e.g., SN2013ej, \citealt{Dhungana+16}).
This transition typically occurs about 100~d post explosion \citep{KasenWoosley09,Sanders+15,Dhungana+16}, consistent with the
timing of our observations relative to the first detection in PSST.

@@

The most effective approach to deal with contaminants like PS15cdi is rapid,
real-time identification. Once a candidate is deemed interesting, optical spectroscopy and
NIR photometry can quickly distinguish between a SN or kilonova/afterglow
origin. If pre-existing templates are not available then the significant aspect is rapid initiation of
follow-up observations at $\lesssim 1$ d that can distinguish the rising phase of a kilonova or
off-axis GRB from a declining SN.

@@

%NOTE THE LACK OF A BIBLIOGRAPHY CALL IN THIS FILE. BIBLIOGRAPHY WORK HAPPENS OUTSIDE THE CHAPTER TEX FILES.

>>>>

\section*{Abstract} 
We report the results of a Dark Energy Camera (DECam)
optical follow-up of the gravitational wave (GW) event GW151226,
discovered by the Advanced LIGO detectors. Our observations cover 28.8
deg$^2$ of the localization region in the $i$ and $z$ bands (containing 3\% 
of the {\tt BAYESTAR} localization probability), starting 10 hours after the 
event was announced and spanning four epochs at $2-24$ days after the 
GW detection. We achieve $5\sigma$ point-source limiting 
magnitudes of $i\approx21.7$ and $z\approx21.5$, with a scatter
of $0.4$~mag, in our difference images. Given the two day delay, 
we search this area for a rapidly declining optical counterpart with
$\gtrsim 3\sigma$ significance steady decline between the first and final
observations. We recover four sources that pass our selection criteria, 
of which three are cataloged AGN. The fourth source is offset
by $5.8$ arcsec from the center of a galaxy at a distance of 187
Mpc, exhibits a rapid decline by $0.5$ mag over $4$ days,
and has a red color of $i-z\approx 0.3$ mag. These properties could satisfy
a set of cuts designed to identify kilonovae. However, this source
was detected several times, starting $94$ days prior to GW151226,  
in the Pan-STARRS Survey for Transients (dubbed as PS15cdi) and
is therefore unrelated to the GW event. Given its long-term 
behavior, PS15cdi is likely a Type IIP supernova that 
transitioned out of its plateau phase during our 
observations, mimicking a kilonova-like behavior. We comment on
the implications of this detection for contamination in future
optical follow-up observations.

@@

The Advanced Laser Interferometer Gravitational-Wave Observatory
(LIGO) is designed to detect the final inspiral and
merger of compact object binaries comprised of neutron stars (NS)
and/or stellar-mass black holes (BH) \citep{LIGOMainRef}. 
The first LIGO observing run (designated O1)
commenced on 18 September 2015 with the ability to detect binary
neutron star (BNS) mergers to an average distance of $\approx 75$ Mpc, a
forty-fold increase in volume relative to the previous generation of
ground-based GW detectors \citep{Martynov+16}. On 2015 September 14 LIGO detected 
the first GW event ever observed, GW150914 \citep{LIGOGW150914}.

The waveform of GW150914 was consistent with the inspiral, merger,
and ring-down of a binary black hole (BBH) system ($36+29$ M$_\odot$; 
\citealt{LIGOGW150914}) providing the first observational
evidence that such systems exist and merge. While there are no robust
theoretical predictions for the expected electromagnetic (EM)
counterparts of such a merger, more than 20 teams conducted a wide
range of follow-up observations spanning from radio to $\gamma$-rays, along with neutrino follow up
\citep{LIGOGW150914FollowUp,LIGOGW150914FollowUpSupp,GW150914IceCube,Annis+16,GW150914Fermi,Evans+16,Kasliwal+16,Savchenko+16,GW150914PS1,GW150914DECam,Tavani+16}.
This effort included deep optical follow-up observations by our group using DECam covering 100 deg$^2$
(corresponding to a contained probability of 38\%~(11\%) of the initial~(final) sky maps)
-- making this one of the most comprehensive optical 
follow-up campaigns for GW150914 \citep{Annis+16,GW150914DECam}. Our search for rapidly
declining transients to limiting magnitudes of $i\approx21.5$ mag
for red $(i-z=1)$ and $i\approx20.1$ mag for blue $(i-z=-1)$ events yielded no counterpart to
GW150914 \citep{GW150914DECam}. One result of the broader multi-wavelength
follow-up campaign is a claimed coincident detection of a weak short
gamma-ray burst (SGRB) from the {\it Fermi}/GBM detector 0.4 s after
the GW event \citep{GW150914Fermi}. However, this event was not
detected in INTEGRAL $\gamma$-ray data \citep{Savchenko+16} and was also
disputed in a re-analysis of the GBM data \citep{Greiner+16}.

A second high-significance GW event, designated GW151226, was
discovered by LIGO on 2015 December 26 at 03:38:53 UT \citep{LIGOGW151226}. 
This event was also due to the inspiral and merger of a BBH system, consisting 
of $14.2^{+8.3}_{-3.7}$~M$_\odot$ and $7.5^{+2.3}_{-2.3}$~M$_\odot$ black holes at a luminosity 
distance of $d_L = 440^{+180}_{-190}$~Mpc \citep{LIGOGW151226}.
The initial localization was provided as a probability sky map via a private 
GCN circular 38 hours after the GW detection \citep{SingerPrice16}. We initiated
optical follow-up observations with DECam 10 hours later on 2015 December 28,
and imaged a 28.8 deg$^2$ region in the $i$ and $z$ bands during several epochs. Here we report 
the results of this search. In \cref{sec:ch4_obs} we discuss the
observations and data analysis procedures. In \cref{sec:ch4_analysis} 
we present our search methodology for potential counterparts to GW151226, 
and the results of this search. We summarize our conclusions in \cref{sec:ch4_conc}. We perform 
cosmological calculations assuming $H_0 = 67.8$~km  s$^{-1}$ Mpc$^{-1}$,
$\Omega_{\lambda} = 0.69$, and $\Omega_m = 0.31$ \citep{Planck2016}. 
Magnitudes are reported in the AB system.

@@

GW151226 was detected on 2015 December 26 at 03:38:53 UT by a 
Compact Binary Coalescence (CBC) search pipeline \citep{LIGOGW151226}. The CBC
pipeline operates by matching the strain data against waveform
templates and is sensitive to mergers containing NS and/or BH. 
The initial sky map was generated by the {\tt BAYESTAR} algorithm
and released 38 hours after the GW detection. {\tt BAYESTAR} is a Bayesian 
algorithm that generates a localization sky map based on the parameter
estimation from the CBC pipeline \citep{Singer+14,SingerPrice16}. The sky area contained within the
initial 50\% and 90\% contours was 430 deg$^2$ and 1340 deg$^2$,
respectively. A sky map generated by the {\tt LALInference}
algorithm, which computes the localization using Bayesian forward-modeling 
of the signal morphology \citep{Veitch+15}, was released on 2016 January 15 UT, 
after our DECam observations had been concluded. The {\tt LALInference} sky map
is slightly narrower than the sky map from {\tt BAYESTAR} with 50\% and 90\% contours 
of 362 deg$^2$ and 1238 deg$^2$, respectively.

We initiated follow-up observations with DECam on 2015 December 28 UT,
two days after the GW detection and 10 hours after distribution of the {\tt BAYESTAR}
sky map. DECam is a wide-field optical imager with a 3 deg$^2$ field of view 
\citep{Flaugher+15}. We imaged a 28.8 deg$^2$ region corresponding to 3\%
of the sky localization probability when convolved with the initial {\tt BAYESTAR} map 
and 2\% of the localization probability in the final {\tt LALInference} sky map. 
The pointings and ordering of the DECam observations were determined using
the automated algorithm described in \citet{GW150914DECam}. The choice 
of observing fields was constrained by weather, instrument availability, and the available
time to observe this sky region given its high airmass. We
obtained four epochs of data with each epoch consisting of one 90 s
exposure in $i$-band and two 90 s exposures in $z$-band for each of the 12 pointings. The first
epoch was obtained 2--3 days after the GW event time (2015
December 28--29 UT), the second epoch was at 6 days (2016 January 1
UT), the third epoch was at 13--14 days (2016 January 8--9), and the
fourth epoch was at 23--24 days (2016 January 18--19). A summary of
the observations is provided in \cref{tab:ch4_datatable} and a visual
representation of the sky region is shown in
\Cref{fig:ch4_obs}.

\begin{deluxetable}{lrccccccc}
\singlespace
\tabletypesize{\footnotesize}
\tablecolumns{9}
\tablewidth{0pt}
\tablecaption{Summary of DECam Observations of GW151226
\label{tab:ch4_datatable}}
\tablehead{	
\colhead{Visit}    &  
\colhead{UT}       & 
\colhead{$\Delta t$\tablenotemark{a}} &
\colhead{$\langle$PSF$_i$$\rangle$}    &
\colhead{$\langle$PSF$_z$$\rangle$}    &
\colhead{$\langle$airmass$\rangle$}  &
\colhead{$\langle$depth$_i\rangle$}   &
\colhead{$\langle$depth$_z\rangle$}  &
\colhead{$A_{\mathrm{eff}}$\tablenotemark{b}} \\
\colhead{ } & 
\colhead{ } &
\colhead{(days)} &
\colhead{(arcsec)} &
\colhead{(arcsec)} &
\colhead{ } &
\colhead{(mag)} &
\colhead{(mag)} &
\colhead{(deg$^2$)} 
}
\startdata
Epoch 1 & 2015-12-28.11 & 1.96 & 0.97 & 0.99 & 1.95 & 22.39 & 22.23 & 14.4 \\      %  6 hexes 
   & 2015-12-29.11 & 2.96 & 1.00 & 0.97 & 1.78 & 22.57 & 22.46 & 14.4 \\      %  6 hexes

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.75 \textwidth]{./figs/chapter4/fig1.pdf}  
\caption{\singlespace
Sky region covered by our DECam observations (red
hexagons) relative to the 50\% and 90\% probability regions 
from the {\tt BAYESTAR} (cyan contours) and {\tt LALInference}
(white contours) localization of GW151226. 
The background color indicates the estimated $5\sigma$ 
point-source limiting magnitude for a 90~s $i$-band exposure 
as a function of sky position for the first night of our DECam 
observations. The variation in the limiting magnitude is largely 
driven by the dust extinction and airmass at that position. 
The dark grey regions indicate sky positions that 
were unobservable due to the telescope pointing limits.  The yellow contour 
indicates the region of sky covered by the Dark Energy Survey 
(DES). The total effective area for the 12 DECam pointings is 28.8 
deg$^2$, corresponding to 3\%~(2\%) of the probability in the
{\tt BAYESTAR}~({\tt LALInference}) sky map.}
\label{fig:ch4_obs}
\end{figure}

@@

\item We require non-negative or consistent with zero 
(i.e., within 2$\sigma$ of zero) 
$i$- and $z$-band fluxes in the difference photometry
across all epochs to eliminate any sources that re-brighten in the 
fourth (template) epoch. This provides an initial sample of 602 candidates.

@@

\clearpage
\item We require a $\ge 3\sigma$ decline in flux between the first and
third epochs to search for significant fading\footnote{\singlespace We note that this 
criterion effectively requires the detection in the first epoch to be $\gtrsim 5\sigma$ 
producing an effectively shallower transient search.  \citet{GW150914DECam} quantified this effect by injecting fake sources into 
their observations to determine the recovery efficiency and loss of detection
depth from analysis cuts. Here, we forego 
such analysis to focus the discussion on the effects of contamination in 
optical follow-up of GW events.}. We calculate $\sigma$ as the 
quadrature sum of the flux errors from the first and third epochs 
($\sigma = \sqrt{\sigma_1^2 + \sigma_3^2}$, where $\sigma_1$ and
$\sigma_3$ are the flux errors from the first and third epochs,
respectively). This criterion leaves a sample of 48 objects.

@@

\item The remaining 32 candidates from step 4 undergo visual inspection. We reject sources
that are present as a point source in the fourth (template) epoch that do not have a galaxy
within 20\arcsec. Sources are cross-checked against NED\footnote{\singlespace \url{https://ned.ipac.caltech.edu/}} 
and SIMBAD\footnote{\singlespace \url{http://simbad.u-strasbg.fr/simbad/}}. This criterion is designed to remove
variable stars and long-timescale transients.
\end{enumerate}

\clearpage
Only four events passed our final criterion. We find that two of those events are 
coincident with the nuclei of known AGN (PKS\,0129-066 and Mrk\,584), indicating 
that they represent AGN variability. A third candidate is coincident with the nucleus of
the bright radio source PMN\,J0203+0956 ($F_\nu(365\,{\rm MHz})\approx 0.4$ Jy, 
\citealt{Douglas+96}), also suggesting AGN variability.

The final candidate in our search is located at RA = 01$^{\rm h}$42$^{\rm m}$16$^{\rm s}$.17 
and DEC = $-$02\arcdeg13\arcmin42 6\arcsec~(J2000), with an offset of $5.8$ arcsec from the galaxy 
CGCG 386-030 (RA =  01$^{\rm h}$42$^{\rm m}$15$^{\rm s}$.6, DEC = $-$02\arcdeg13\arcmin38 
5\arcsec; J2000), at $z = 0.041$ or $d_L \approx 187$ Mpc (6dFGS, \citealt{Jones+04,Jones+09}); see 
\Cref{fig:ch4_PS15cdi}. We note that this distance is inconsistent with the 90\% confidence interval for
the distance to GW151226 based on the GW data \citep{LIGOGW151226}. We observe this source in a state of rapid decline with an 
absolute magnitude of $M_i \approx -15$~mag on 2015 December 28 and $M_i\approx -14.5$~mag on 2016 
January 1, indicating a decline rate of $\approx 0.12$ mag d$^{-1}$; the decline rate in $z$-band is 
$\approx 0.10$ mag d$^{-1}$. Additionally, the source exhibits a red $i-z$ color of $0.3$ mag.

@@

\begin{figure}[t!]
\centering
	\includegraphics[width=0.9\textwidth]{./figs/chapter4/fig2.pdf}
\caption{\singlespace {\it Top}: Single-epoch images of our main candidate
from all four epochs (green circle). This is the event discovered as PS15cdi in
the PSST about 94 d prior to GW151226. {\it Bottom}: Light
curve data for PS15cdi from PSST $w$- and $i$-band
observations (green squares and yellow diamonds, respectively).
Our DECam $i$- and $z$-band data are shown as
blue circles and red stars, respectively. The revised DECam analysis 
using pre-existing templates is shown as open symbols. Upper
limits are indicated by triangles. The inset focuses on our
DECam data, indicating a rapid decline in both $i$ and $z$
bands.  We fit a power-law model to the data finding a temporal index of 
$\alpha = -0.43$ (dashed-dot line). Kilonova models from \citet{BarnesKasen13} with $v_{\rm ej}=0.2c$ and $M_{\rm ej}=0.1$ M$_\odot$ 
at a distance of $187$~Mpc are also shown (dashed line).}
\label{fig:ch4_PS15cdi}
\end{figure}

This source was previously detected as PS15cdi on 2015 September 23 by the Pan-STARRS 
Survey for Transients (PSST\footnote{\singlespace \url{http://psweb.mp.qub.ac.uk/ps1threepi/psdb/candidate/1014216170021342600/}},
\citealt{Huber+15}); see \Cref{fig:ch4_PS15cdi}. The absolute $i$-band 
magnitude in the first PSST epoch, $M_i\approx -16.6$ mag and the shallow decline of $\approx0.6$~mag 
over $\approx70$~d, are consistent with a Type IIP core-collapse supernova (SN). 
A likely interpretation of the rapid decline in our observations is that PS15cdi is a Type IIP SN undergoing 
the rapid transition from the hydrogen recombination driven plateau to the radioactive $^{56}$Co dominated 
phase \citep{KasenWoosley09,Sanders+15,Dhungana+16}. The red $i-z$ color in our data is consistent with observations 
of other IIP SN during this phase of evolution (e.g., SN2013ej, \citealt{Dhungana+16}).
This transition typically occurs about 100~d post explosion \citep{KasenWoosley09,Sanders+15,Dhungana+16}, consistent with the 
timing of our observations relative to the first detection in PSST.

@@

\clearpage
The most effective approach to deal with contaminants like PS15cdi is rapid,
real-time identification. Once a candidate is deemed interesting, optical spectroscopy and 
NIR photometry can quickly distinguish between a SN or kilonova/afterglow
origin. Specifically, the kilonova spectrum will be redder, with clear suppression below $\sim 6000$~\AA~due
to the opacities of $r$-process elements \citep{Kasen+13}. By comparison, the SN spectrum will appear bluer and
dominated by iron group opacities \citep{Kasen+13}, while the afterglow spectrum will exhibit a featureless power-law
spectrum \citep{Berger2014}. If pre-existing templates are not available then the significant aspect is rapid initiation of
follow-up observations at $\lesssim 1$ d that can distinguish the rising phase of a kilonova or 
off-axis GRB from a declining SN.

@@
